---
title: "Analyzing Key Factors of Calorie Burn Efficiency in a Workout"
editor: visual
---

## Motivation and Context

Often times, we do our research on calories involving diet, exercise, and lifestyle overall because calories are the energy we need to function throughout the day. We burn calories in order to balance our calorie intake, improve metabolic efficiency, and reduce stress. However, with busy daily routines, finding time for this can prove to be difficult.

In this investigation, I will analyze the demographics and living habits of users in order to investigate any correlations with calorie burn in a workout.

I was pretty active in middle school and high school, but once I entered college and started working, finding time for physical activity grew more difficult. I tried many different types of workout routines and eventually developed one tailored to me I enjoy and fits in my busy school and work schedule. I noticed that when I worked out pretty consistently, I got better sleep and my mood improved, which sparked my interest in the topic.

## Main Objective

The data set is designed to help analyze and predict workout efficiency. However because only one workout session for each user was recorded, I decided to analyze what variables are more effective in burning calories during a workout.

## Packages Used In This Analysis

```{r}
#| label: load packages 
#| message: false 
#| warning: false  

library(here) 
library(readr) 
library(dplyr) 
library(ggplot2) 
library(rsample)  
library(tidyverse)
library(tidymodels)
library(probably)
```

| Package | Use |
|------------------------------------|------------------------------------|
| [here](https://github.com/jennybc/here_here) | to easily load and save data |
| [readr](https://readr.tidyverse.org/) | to import the CSV file data |
| [dplyr](https://dplyr.tidyverse.org/) | to massage and summarize data |
| [ggplot2](https://ggplot2.tidyverse.org/) | to create nice-looking and informative graphs |
| [rsample](https://rsample.tidymodels.org/) | to split data into training and test sets |
| [tidyverse](https://www.tidyverse.org/) | to simplify data manipulation, designed for data science |
| [tidymodels](https://www.tidymodels.org/) | to use modeling and machine learning |
| [probably](https://probably.tidymodels.org/) | to investigate and estimate optimal probability thresholds |

## Data Description

Our data set contains fitness data on user's

:::::: columns
::: {.column width="33%"}
**Demographic and Body**

-   Age

-   Gender

-   Height (centimeters)

-   Weight (kilograms)

-   Body Fat (percentage)

-   VO₂ Max

-   Resting Heart Rate
:::

::: {.column width="33%"}
**Lifestyle and Wellness**

-   Sleep Hours

-   Water Intake (liters)

-   Daily Calories Intake

-   Mood Before Workout

-   Mood After Workout
:::

::: {.column width="33%"}
**Workout**

-   Workout Type

-   Workout Intensity

-   Workout Duration (minutes)

-   Distance (kilometers)

-   Steps Taken

-   Heart Rate (bpm)

-   Calories Burned
:::
::::::

This data set is from Adil Shamim from Kaggle and synthetically generated to simulate real-world fitness tracking data. The data mimics information collected from different fitness apps.

```{r}
#| label: Loading Data 
#| echo: false
#| results: 'hide'
fitness <- read_csv("workout_fitness_tracker_data.csv", show_col_types = FALSE)
```

### Data Limitations

As mentioned earlier, this data is synthetically generated and only represents similar real user data and some relationships between variables may not be fully precise. On top of that, the creater of the data set used Python to create randomly generated numbers in a given range for each variable, without considering real-world characteristics they could have. For example, most people range between a height of 159cm-171cm, which should mean our height variable should have a normal distribution. However, when we take a look at the actual variable height, we see that there are an equal frequency within the range of 150cm and 199cm, which could potentially cause bias.

Another limitation in this data set is the potential of multicollinearity in the data, which can affect the model's performance.

## Data Wrangling

```{r}
#| label: Data Wrangling
#| echo: false
#| results: 'hide' 
# Checking for NAs 
sum(is.na(fitness))  
```

I checked the data set for any missing or NA variables and split the data into a training (75%) and testing set (25%). Before I chose what I was going to analyze, I manipulated the variables in a way I thought was fit for this study and ended up not needing to do this, but felt it was important to include as a part of my data wrangling.

::::: columns
::: column
**BMI**

I converted the height into meters and took the weight and divided by the height (meters) squared to get the user's BMI. These values were then sorted into:

-   BMI \< 18.5, "Underweight"

-   18.5 \< BMI \< 24.9, "Normal Weight"

-   24.9 \< BMI, "Overweight"
:::

::: column
**Mood Change**

I converted the categorical variable Mood Before Workout and Mood After Workout into a numerical variable:

-   Happy or Energized= 2

-   Neutral = 1

-   Stressed/Tired or Fatigued= 0

and then took the difference between the values. If the value was positive, mood improved; If the value was negative, mood worsened.
:::
:::::

```{r, eval=FALSE, include=FALSE}
#| label: Changing and Mutating Variables 
# Changing variables 
fitness$`Mood Before Workout` <- ifelse(fitness$`Mood Before Workout` == "Happy", 2,
                                        ifelse(fitness$`Mood Before Workout` == "Neutral", 1,
                                               ifelse(fitness$`Mood Before Workout` == "Stressed", 0,
                                                      ifelse(fitness$`Mood Before Workout` == "Tired", 0, NA))))

fitness$`Mood After Workout` <- ifelse(fitness$`Mood After Workout` == "Energized", 2,                      
                                       ifelse(fitness$`Mood After Workout` == "Neutral", 1,                     
                                              ifelse(fitness$`Mood After Workout` == "Fatigued", 0, NA)))  

# Mutate 
fitness <- fitness |>   
  mutate(
    mood_change = `Mood After Workout` - `Mood Before Workout`,     
    bmi = `Weight (kg)`/(`Height (cm)`/100)^2,     
    bmi = if_else(bmi < 18.5, "Underweight",
                  ifelse(bmi < 24.9, "Normal Weight",       
                         ifelse(bmi < 29.9, "Overweight", "Obesity")))
    )
```

```{r}
#| label: Training and Testing Sets 
#| echo: false
#| results: 'hide' 
fitness_split <- initial_split(fitness, prop = 0.75) 
fitness_train <- training(fitness_split) 
fitness_test <- testing(fitness_split)
```

## Exploritory Data Analysis

When examining how each variable relates to Calories Burned, most exhibit similar patterns. Below is a comparison highlighting how both a categorical and a numerical variable are associated with Calories Burned.

```{r}
#| label: EDA Categorical
#| echo: false
# Looking at categorical variables and getting a general idea of the behavior with calories burned
ggplot(data = fitness, 
       mapping = aes(x = `Calories Burned`)) +
  geom_histogram(fill = "mediumpurple1", 
                 color = "black",
                 center = 500,
                 binwidth = 200) +
  labs(x = "Calories Burned",
       y = "Workout Type") +
  facet_wrap(vars(`Workout Type`))
```

Here we see the frequency of calories burned for each type of workout. Almost all of the graphs look similar in frequency, which demonstrates the limitation of this data set.

```{r}
#| label: EDA Numerical
#| echo: false
# Looking at numerical variables and getting a general idea of the behavior with calories burned
ggplot(data = fitness, 
       mapping = aes(x = `Workout Duration (mins)`, 
                     y = `Calories Burned`)) +
  geom_point(color = "mediumpurple1", size = 2) +
  labs(x = "Workout Duration (Minutes)",
       y = "Calories Burned")
```

Even more concerning, we observe that some workouts burn a significant number of calories in a short amount of time, while others yield only a small calorie burn despite extended durations.

## Modeling

```{r}
#| label: Finding Significant Variables through LASSO 
#| echo: false
#| warning: false  
#| results: 'hide'
lasso_model <- linear_reg(mode = "regression", engine = "glmnet",
                          penalty = tune(),                           
                          mixture = 1)

lasso_recipe <- recipe(`Calories Burned` ~ Age + Gender + `Height (cm)` + `Weight (kg)` + `Workout Type` + 
                       `Workout Duration (mins)` + `Heart Rate (bpm)`+ `Steps Taken` + `Distance (km)` + 
                       `Workout Intensity` + `Sleep Hours` + `Water Intake (liters)` + `Daily Calories Intake` + 
                       `Resting Heart Rate (bpm)` + `VO2 Max` + `Body Fat (%)` + `Mood Before Workout` + 
                       `Mood After Workout`,   data = fitness_train) |>   
  step_normalize(all_numeric_predictors()) |> # don't scale the response   
  step_dummy(all_nominal_predictors())  

lasso_wflow <- workflow() |>   
  add_model(lasso_model) |>   
  add_recipe(lasso_recipe)  

fitness_cv <- vfold_cv(fitness_train, v = 10)  

lasso_tune <- tune_grid(lasso_model,
                        lasso_recipe,                          
                        resamples = fitness_cv,                          
                        grid = grid_regular(penalty(range = c(-5, 0)), levels = 50))  

lasso_best <- lasso_tune |>   
  select_by_one_std_err(     
    metric = "rmse",     
    desc(penalty) ) 

lasso_wflow_final <- lasso_wflow |>   
  finalize_workflow(parameters = lasso_best) 

lasso_pred_check <- lasso_wflow_final |>   
 fit_resamples(     
   resamples = fitness_cv,     
   control = control_resamples(save_pred = TRUE) ) |>   
 collect_predictions()  

lasso_fit <- lasso_wflow_final |>   
  fit(data = fitness_train)  

lasso_coef <- lasso_fit |>   
  broom::tidy(penalty = lasso_best$penalty)
```

I chose to use obtain a LASSO model with the data set because I wanted to see which variables were most significant in burning calories. As a result, I successfully identified that the higher a user's age, steps taken, sleep hours, daily calories intake, resting heart rate, mood after workout, mood change, gender, and workout type contributed to burning more calories. Also, the lower a user's height, weight, workout duration and heart rate contributed to burning more calories. Because Calories Burned is a numerical variable, I chose to select the model using a RMSE. RMSE represents how far off I am in calories.

```{r}
#| label: Model Selection 
#| echo: false
selected_model <- linear_reg(mode = "regression", engine = "lm") 

selected_recipe <- recipe(`Calories Burned` ~ Age + Gender + `Height (cm)` + `Weight (kg)` + `Workout Type` +                              `Workout Duration (mins)` + `Heart Rate (bpm)`+ `Steps Taken` + `Workout Intensity` + 
                          `Sleep Hours` + `Daily Calories Intake` + `Resting Heart Rate (bpm)` + 
                          `Mood After Workout`, data = fitness_train)  

selected_wflow <- workflow() |>   
  add_model(selected_model) |>   
  add_recipe(selected_recipe)  

selected_lm <- lm(`Calories Burned` ~ Age + `Height (cm)` + `Weight (kg)` + `Workout Duration (mins)` + 
                  `Heart Rate (bpm)`+ `Steps Taken` + `Sleep Hours` + `Daily Calories Intake` + 
                  `Resting Heart Rate (bpm)`, data = fitness_test)  

summary(selected_lm)
```

## Insights

Despite using LASSO regression for variable selection, no predictors were significantly important under sample splitting. This suggests some multicollinearity within the variables.

```{r}
#| label: RMSE across lambda
#| echo: false
lasso_tune |>
  collect_metrics() |>
  filter(.metric == "rmse") |>
  ggplot(mapping = aes(x = penalty, y = mean)) + geom_point() + geom_line() +
  scale_x_log10()
```

In this figure above, we see that the trend has a huge decrease as penalty grows. This suggests that more variables improves performance, but not by much, which further explains our LASSO model.

```{r}
#| label: LASSO Coefficient Path
#| echo: false
lasso_fit |>
  extract_fit_engine() |>
  plot(xvar = "lambda", label = TRUE)
```

This figure shows the LASSO coefficient path, each line representing a predictor's coefficient. We can see that no variables remained dominant as penalty increased. Again suggesting little significance.

### Limitations and Future Work

The model is poorly predicting the outcome of our response variable due to the synthetic data. Because of this, the data lacks the complexity and meaningful relationships typically found in actual fitness data. Future improvements could be the use of an actual fitness data set with more features that might reveal stronger relationships. Also maybe using nonlinear methods that can reveal what linear methods missed because LASSO assumes a linear relationship and well as independence.

There are no privacy concerns involving this data set because these numbers were randomly generated, however had we used real data, we would need to consider data privacy and user transparency.

## Citations

<https://www.kaggle.com/datasets/adilshamim8/workout-and-fitness-tracker-data>
